---
title: "Expt2. data preprocessing and initial analysis"
author: "Kent Riemondy RBI"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: false
    theme: cosmo
    highlight: tango
    fig_caption: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      cache = TRUE)
```


```{r libs, cache = F}
source(here::here("R", "utils.R"))

walk(c(fig_dir, mkrs_dir, tbls_dir, obj_dir),
     dir.create, showWarnings = F)
seed_value <- 20210413
```

# Experiment Summary

Experimental description taken from grant application

Background Chronic obstructive pulmonary disease (COPD) is a devastating disease and the third leading cause of death worldwide. COPD is characterized by progressive and devastating loss of alveolar tissue. Endogenous lung stem cells and tissue repair pathways are thought be dysfunctional and their reactivation is a promising therapeutic approach for COPD. Wnt/β-catenin signaling is critical for stem cell function in many organs and is reduced in human COPD lung tissue. We have previously demonstrated that therapeutic Wnt/-catenin activation can initiate tissue repair in COPD patient-derived lung tissue ex vivo. Recently, we identified a distal lung epithelial stem cell, club cell (~5% of whole lung epithelium), which showed impaired stem cell function in a well-established elastase induced murine COPD model in our laboratory. Importantly, these cells could be re-activated to regenerate alveolar cells by Wnt/β-catenin signal activation in vitro. Thus, we hypothesize that club cells represent a novel Wnt-responsive epithelial stem cell population and that their dysfunction may drive COPD pathogenesis. To date, however, the functions of club cells and the role of Wnt/β-catenin and other signaling pathways during dysregulation of club cells in COPD, remain elusive.

## Approach (taken from proposal)

To identify genes and signaling pathways regulating the dysfunction of club cells in COPD, we have recently established an inducible double labeling mouse model (Scgb1a1-CreERT;Rosa-dtTomato;TCF-GFP), which allows lineage tracing of club cells by dtTomato expression as well as real-time labeling of the Wnt/β-catenin activity with EGFP expression in these cells. Here, we propose to treat this mouse line with elastase and saline to generate COPD and control mice, respectively, followed by isolation of tdTomato labeled club cells for single cell RNA sequencing analysis. 

In the second experiment, there are 2 sample per treatment, captured using the 10X Genomics 3’ Platform and subsequent sequencing.


```{r get_data}
data_dir <- file.path(proj_dir, "data", "2021-03-11")

samples <- c(
  "E1",
  "E2",
  "S1",
  "S2"
)

sample_paths <- file.path(data_dir,
                          samples,
                          "outs", 
                          "filtered_feature_bc_matrix")

names(sample_paths) <- samples
mat <- Read10X(sample_paths)
```

## General QC {.tabset}

```{r}
sample_paths <- file.path(data_dir,
                          samples,
                          "outs", 
                          "raw_feature_bc_matrix")

names(sample_paths) <- samples

barcode_counts <- imap_dfr(sample_paths, 
               ~{
                 cnts <- read10xCounts(.x, col.names = TRUE) %>% counts(.) 
                 tibble(
                   cell = colnames(cnts),
                   cell_id = str_c(.y, "_", cell),
                   nUMI = Matrix::colSums(cnts))},
               .id = "sample") %>% 
  filter(nUMI > 0) %>% 
  arrange(desc(nUMI)) %>% 
  mutate(is_cell = cell_id %in% colnames(mat)) %>% 
  split(., .$sample) %>% 
  map(~mutate(.x, cell_rank = row_number()))
                        
p <- imap(barcode_counts,
    ~ggplot(.x, aes(cell_rank, nUMI)) +
      geom_point(aes(color = is_cell)) +
      annotate("text", x = 10, y = 5, 
               label = str_c("n = ", scales::comma(sum(.x$is_cell)))) + 
      scale_color_manual(values = palette_OkabeIto, 
                         name = "Is Cell?") + 
      labs(title = .y,
           x = "Barcode Rank",
           y = "# of UMIs") +
      scale_x_log10() + 
      scale_y_log10())

```

 
```{r, results ='asis'}

for(i in seq_along(p)){
  cat('\n### Barcode Plot Sample: ', names(barcode_counts)[i], '\n')
  print(p[[i]])
  cat('\n')
}
```

### Other library prep metrics
```{r, fig.width = 12, fig.height=18}
metrics_paths <- file.path(data_dir,
                          samples,
                          "outs", 
                          "metrics_summary.csv")

names(metrics_paths) <- samples

mapping_dat <- map_dfr(metrics_paths, read_csv, .id = "sample", 
                       col_types = cols(
                         `Estimated Number of Cells` = col_number(),
                         `Mean Reads per Cell` = col_number(),
                         `Median Genes per Cell` = col_number(),
                         `Number of Reads` = col_number(),
                         `Valid Barcodes` = col_character(),
                         `Sequencing Saturation` = col_character(),
                         `Q30 Bases in Barcode` = col_character(),
                         `Q30 Bases in RNA Read` = col_character(),
                         `Q30 Bases in UMI` = col_character(),
                         `Reads Mapped to Genome` = col_character(),
                         `Reads Mapped Confidently to Genome` = col_character(),
                         `Reads Mapped Confidently to Intergenic Regions` = col_character(),
                         `Reads Mapped Confidently to Intronic Regions` = col_character(),
                         `Reads Mapped Confidently to Exonic Regions` = col_character(),
                         `Reads Mapped Confidently to Transcriptome` = col_character(),
                         `Reads Mapped Antisense to Gene` = col_character(),
                         `Fraction Reads in Cells` = col_character(),
                         `Total Genes Detected` = col_number(),
                         `Median UMI Counts per Cell` = col_number()
                       ))

clean_up_metadata <- function(metrics_summary) {
  metrics_summary <- mutate_all(metrics_summary, str_remove, "%$")
  metrics_summary <- mutate_at(metrics_summary, .vars= 2:ncol(metrics_summary), as.numeric)
  metrics_summary
}

mapping_dat <- clean_up_metadata(mapping_dat)

metrics <- colnames(mapping_dat)[2:ncol(mapping_dat)]
mapping_dat <- pivot_longer(mapping_dat, 
                            cols = -sample, 
                            names_to = "metric") %>% 
  mutate(metric = str_wrap(metric, 20))

p <- ggplot(mapping_dat, aes(sample, value)) +
  geom_col(aes(fill = sample)) +
  facet_wrap(~metric, 
             scales = "free",
             ncol = 4) + 
  scale_fill_brewer(palette = "Paired") + 
  labs(x = "", y = "") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "none") 

p 

save_plot(file.path(fig_dir, "qc_metrics.pdf"),
          p,
          nrow = 5,
          ncol = 4,
          base_asp = 1)
```


```{r create_seurat, message = F, results = 'hide', warning = F}
transgenes <- c("tdtomato-sv40", "egfp-sv40")

so <- CreateSeuratObject(
  mat,
  min.cells = 10,
  min.features = 200,
  names.delim = "_",
  names.field = 1
)

rm(mat) ; gc()

so <- PercentageFeatureSet(so, 
                             pattern = "^mt-", 
                             col.name = "percent.mt")

so@meta.data$sample <- so@meta.data$orig.ident
so@meta.data$treatment <- ifelse(str_detect(so@meta.data$orig.ident, "^E"),
                              "Elastase",
                              "Saline")

so@meta.data$treatment <- factor(so@meta.data$treatment, levels = c("Saline", "Elastase"))

```

### Percent Mitochondrial UMIs 

```{r, fig.cap = "The low mitochondrial UMI % suggests high quality cells"}


p <- plot_violin(so@meta.data, 
            "sample",
            "percent.mt",
            .fill = "treatment",
            cols = scale_fill_manual(values = palette_OkabeIto)) +
  labs(x = "", y = "% UMIs from Mitochondrial RNAS") 

save_plot(file.path(fig_dir, "prop_mito.pdf"),
          p, base_asp = 1.75)
p
```


### # of genes detected 

```{r}
p <- plot_violin(so@meta.data,
            "sample", 
            "nFeature_RNA",
            .fill = "treatment",
            cols = scale_fill_manual(values = palette_OkabeIto)) +
  labs(x = "", y = "# of genes per cell")

save_plot(file.path(fig_dir, "ngene.pdf"),
          p, base_asp = 1.75)
p
```


### # of UMIs detected

```{r}
p <- plot_violin(so@meta.data, 
            "sample",
            "nCount_RNA", 
            .fill = "treatment",
            cols = scale_fill_manual(values = palette_OkabeIto)) +
  labs(x = "", y = "# of UMIs") + 
        scale_y_continuous(breaks = scales::pretty_breaks(n = 10))

save_plot(file.path(fig_dir, "numi.pdf"),
          p, base_asp = 1.75)
p
```

### Table of mitochondrial proportions per sample

```{r, message = FALSE}
so@meta.data %>% 
  group_by(sample) %>% 
  summarize(median_percent_mito = median(percent.mt), 
            mean_percent_mito = mean(percent.mt)) %>% 
  arrange(desc(median_percent_mito))
```


### Relationship between UMIs and % mitochondria {.tabset}

```{r}
so$outlier_percent_mt <- check_outliers(so$percent.mt, 
                                        so$sample, 
                                        min = 5, 
                                        max = 20)

sample_names <- as.character(unique(so@meta.data$orig.ident))
per_sample <- map(sample_names, ~filter(so@meta.data, 
                                        orig.ident == .x))
p <- list()
for(i in seq_along(per_sample)){
  .col <- discrete_palette_default[i]
  p[[i]] <- ggplot(per_sample[[i]], aes(nCount_RNA, percent.mt)) +
        geom_point(aes(color = outlier_percent_mt)) +
        scale_color_manual(values = palette_OkabeIto) +
        scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
        scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
    theme(axis.text.x = element_text(angle= 90, hjust = 1, vjust = 0.5)) +
    labs(subtitle = unique(per_sample[[i]]$orig.ident))
}

plt <- plot_grid(plotlist = p, nrow = 2, ncol = 2)

save_plot(file.path(fig_dir, "umi_v_prop_mito.pdf"),
          plt,
          base_asp = 1, 
          nrow = 2,
          ncol = 2)

plt
```


## Filter cells and samples.


Overall the sample quality looks quite excellent, with a good # of cells captured, low # of cells with high mitochondrial reads, and low background from lysed cells.

Suggest using some minor filters to remove cells with outlier mitochondrial UMIs (>3 MAD or > 20%), > 20K UMI or < 500 UMIs, > 5000 genes detected . 

```{r rows.print = 30}

prev_count <- dplyr::count(so@meta.data, sample)
filter_summary <- list(
  "Cells called by cellranger" = prev_count)

so <- subset(so, subset = outlier_percent_mt, invert = T)
cur_count <- dplyr::count(so@meta.data, sample)
filter_summary$"Too high Mitochondrial reads" <- cur_count %>%
  left_join(prev_count, by = "sample") %>% 
  mutate(n = n.y - n.x) %>% 
  dplyr::select(sample, n)
prev_count <- cur_count

so <- subset(so, subset = nCount_RNA < 20000)
cur_count <- dplyr::count(so@meta.data, sample)
filter_summary$"Too high # of UMIs" <- cur_count %>%
  left_join(prev_count, by = "sample") %>% 
  mutate(n = n.y - n.x) %>% 
  dplyr::select(sample, n)
prev_count <- cur_count

so <- subset(so, subset = nFeature_RNA < 5000)
cur_count <- dplyr::count(so@meta.data, sample)
filter_summary$"Too low # of genes" <- cur_count %>%
  left_join(prev_count, by = "sample") %>% 
  mutate(n = n.y - n.x) %>% 
  dplyr::select(sample, n)
prev_count <- cur_count

so <- subset(so, subset = nCount_RNA > 500)
cur_count <- dplyr::count(so@meta.data, sample)
filter_summary$"Too low # of UMIs" <- cur_count %>%
  left_join(prev_count, by = "sample") %>% 
  mutate(n = n.y - n.x) %>% 
  dplyr::select(sample, n)
prev_count <- cur_count

filter_summary$"Passed QC" <- cur_count

cell_counts_post_qc <- bind_rows(filter_summary, .id = "Catagory") %>% 
  dplyr::rename("# of cells" = n,
                Category = Catagory)
cell_counts_post_qc %>% 
  arrange(Category)
```

```{r, fig.cap = "Cell counts pre/post filtering."}
filter(cell_counts_post_qc, Catagory != "Cells called by cellranger") %>% 

ggplot(aes(sample, `# of cells`)) +
  geom_col(aes(fill = Catagory)) + 
  scale_fill_manual(values = discrete_palette_default,
                    name = "") +
  labs(x = "sample",
       y = "# of cells")
```

## Normalize and embed into 2D with UMAP

```{r}
opts <- list(
  n_var = 2000,
  ndims = 20,
  k = 15,
  min_dist = 0.3,
  coarse_clustering = "RNA_snn_res.0.1",
  refined_clustering = "RNA_snn_res.1"
)
```

For this dataset mostly standard Seurat processing will be performed. The top `r opts$n_var` variable genes will be used for PCA. The top `r opts$ndims` PCs will be used to generate a UMAP projection (min_dist = `r opts$min_dist`) and to perform clustering at various resolutions. Two clustering settings will be used to capture broad or more refined clusters. 

```{r, message=FALSE, warning=FALSE}
so <- NormalizeData(so)

so <- FindVariableFeatures(so,
                           selection.method = "vst",
                           nfeatures = opts$n_var,
                           verbose = FALSE)

# remove transgenes from variable features 
var_info <- so@assays$RNA@meta.features
var_genes <- rownames(var_info)
var_info <- var_info[!var_genes %in% transgenes, ]
var_info$vst.variable <- FALSE
var_info[order(-var_info$vst.variance.standardized)[1:opts$n_var], ]$vst.variable <- TRUE
so@assays$RNA@meta.features <- var_info
so@assays$RNA@var.features <- rownames(var_info)[var_info$vst.variable]
  
so <- ScaleData(so, verbose = FALSE)

so <- RunPCA(so, 
             npcs = 40, 
             verbose = FALSE, 
             seed.use = seed_value)

so <- FindNeighbors(so, 
                    reduction = "pca", 
                    dims = 1:opts$ndims, 
                    k.param = opts$k, 
                    verbose = FALSE)

so <- RunUMAP(so, 
              dims = 1:opts$ndims,
              n.neighbors = opts$k,
              min.dist = opts$min_dist, 
              seed.use = seed_value,
               verbose = FALSE)

so <- FindClusters(so, 
                   resolution = seq(0.1, 1, 0.1),
                   random.seed = seed_value,
                    verbose = FALSE)
```

### by treatment

```{r}
plot_umap(so, "sample")
```


```{r, fig.cap = "Overall the samples overlap for most populations, however Saline 1 and 2 show distinct populations in a large cluster (which is AT2 cells)."}
plot_umap(so, "treatment", group = "sample", .cols = palette_OkabeIto)
```

### by cluster

```{r, warning = FALSE, fig.height=16, fig.width=16, fig.cap = "Clustering at various resolution settings." }
cluster_cols <- str_subset(colnames(so@meta.data), "RNA_snn")

plot_umap(so, 
          cluster_cols, 
          label_text = TRUE, 
          label_color = "black",
          ggrepel_opts = list(force = 0)) %>% 
  plot_grid(plotlist = .,
            nrow = 4,
            ncol = 3)
```


Two clustering settings will be used; `r opts$coarse_clustering` as the "coarse" clustering and 
`r opts$refined_clustering` as a refined clustering setting to examine sub-states within cell types. 

```{r}
so$coarse_clusters <- so@meta.data[[opts$coarse_clustering]] 
so$refined_clusters <- so@meta.data[[opts$refined_clustering]]
Idents(so) <- "refined_clusters"
```

```{r}
plts <- plot_umap(so, c("coarse_clusters", 
                "refined_clusters"), 
                group = "sample",
                sorted = "random")

names(plts) <- c("coarse_clusters", 
                "refined_clusters")

```

```{r, results ='asis'}
for(i in seq_along(plts)){
  cat('\n### UMAP by ', names(plts)[i], '\n')
  print(plts[[i]])
  cat('\n')
}
```


## Annotate cell types

To annotate cell types we use use the `clustifyr` package developed by Rui Fu in the RBI. We will compare the refined clusters defined above to a reference dataset of cell types. The reference data will be taken from the `Strunz et al. Nature Commmunications 2020` paper in which whole lung or lung epithelia were profiled with single cell RNA-seq. The paper used a bleomycin injury model, but for this analysis only control  (PBS) non-injured cells will be used to generate reference cell types. Note however that there are some control cells annotated as the injured cell states, "e.g. activate club cells", so these cell states may also be present. 


```{r, fig.width = 9, fig.height=9, fig.cap = "Heatmap of correlation coefficients between the refined clusters in our data and the reference cell types defined in the Strunz et al dataset.", message = FALSE}
library(clustifyr)
library(ComplexHeatmap)
ref <- readRDS(here("results", "objects", "atlas", "ctrl_reference.rds"))

res <- clustify(so, 
                cluster_col = "refined_clusters",
                query_genes = VariableFeatures(so),
                ref_mat = ref, 
                obj_out = FALSE,
                compute_method = "pearson") 

hm <- Heatmap(t(res),
              col = viridis::magma(256),
              row_title = "Cell types (Strunz et al)",
              column_title_side = "bottom",
              column_title = "clusters",
              name = "Pearson")

pdf(file.path(fig_dir, "cell_type_cor_heatmap.pdf"),
    width = 9,
    height = 9)
draw(hm)
a <- dev.off(); rm(a)

hm
```


Using these reference datasets as a guide, each cluster will be assigned to the most correlated cell type in the reference. Most clusters have very high correlation to the cell types in the reference lung dataset, as expected. 

```{r, rows.print = 15, fig.cap = "Cell type assignment based on highest correlation values."}
cor_to_call(res) %>% 
  dplyr::rename(cell_type = type,
         cor_val = r)
```

### Annotated cell types

```{r fig.cap = "UMAP of annotated cell types ", message = FALSE}
so <- clustify(so, 
                cluster_col = "refined_clusters",
                query_genes = VariableFeatures(so),
                ref_mat = ref, 
                obj_out = TRUE,
                compute_method = "pearson") 
so@meta.data$UMAP_1 <- NULL
so@meta.data$UMAP_2 <- NULL
so@meta.data$r <- NULL
so@meta.data$coarse_cell_type <- so@meta.data$type
so@meta.data$type <- NULL

so@meta.data$coarse_cell_type <- str_split(so@meta.data$coarse_cell_type, "_") %>% 
    map_chr(~.x[length(.x)]) %>% 
  str_to_title()

p <- plot_umap(so,
  "coarse_cell_type",
  group = "treatment",
  legend_title = "",
  .cols = discrete_palette_default
)

save_plot(file.path(fig_dir, "umap_of_annotated_cell_types.pdf"),
          p,
          base_asp = 2)
p
```

## Check on doublets

The club cells appear to have a subset of cells that overlap with the AT2 cell populations, which is suspect and may be potential doublets. I'll examine doublets using an algorithm that mixes cells in silico to simulate doublets. This is done per sample  however the results are shown in aggregate for simplicity.

```{r, fig.width = 10}
so <- score_doublets(so, cluster_col = "refined_clusters", sample_col = "sample")

p1 <- plot_violins(so, "refined_clusters", "doublet_score")
p2 <- plot_umap(so, "doublet_score")

plot_grid(p1, p2, nrow = 1)
```

Based on this analysis we will exclude cluster 13, however the at2-like club cells don't appear to be doublets. 

```{r}
so <- subset(so, subset = refined_clusters == "13", invert = TRUE)
```

### Annotated cell types with subclusters

The refined cluster annotations are merged with the cell type labels and reordered so that each subpopulation starts with 1 for each cell type. 

```{r, fig.cap = "UMAP of cell types with  subclusters also annotated"}

so@meta.data$cell_type_cluster <- add_subcluster_ids(so, "coarse_cell_type", "refined_clusters")
p <- plot_umap(so, "cell_type_cluster", 
          group = "sample",
          legend_title = "")

save_plot(file.path(fig_dir, "umap_of_annotated_cell_types_with_clusters.pdf"),
          p,
          base_asp = 2)
p

```

### Show known cell type markers

```{r}
mkr_list <- read_excel(here("dbases", "mouse_lung_epithelial_cell_marker_list.xlsx")) %>% 
  split(., 1:nrow(.)) 
  
last_seen <- NA
grps <- vector("character", length(mkr_list))
for (i in seq_along(mkr_list)){
  val <- mkr_list[[i]]$List
  if(is.na(val)){
      grps[i] <- last_seen
  } else {
      last_seen <- val
      grps[i] <- val
  }
}
mkr_list <- bind_rows(mkr_list) %>% 
  mutate(cell_type = grps) %>% 
  dplyr::select(-List) %>% 
  split(., .$cell_type) %>% 
  map(~.x$Name)

mkr_list <- c(mkr_list, 
              list("Transgenes" = transgenes)
              )

mkr_list <- map(mkr_list, ~.x[.x %in% rownames(so)])
```

```{r, fig.width = 10, fig.height = 24, fig.cap = "Montage of UMAPs of cell types markers"}
plot_montage <- function(sobj,
                         feature_list,
                         ncols = 6,
                         outfile = NULL,
                         plot_fxn = plot_umap,
                         plot_args = list(minimal_theme = TRUE,
                                          sorted = "random")) {
  
  all_plts <- list()
  rel_heights <- c()
  for(i in seq_along(feature_list)){
    
    features <- feature_list[[i]]
    id <- names(feature_list)[i]
    n_plts <- length(features)
    
    plts <- do.call(plot_fxn, args = c(list(seurat_obj = sobj, 
                                            feature = features), 
                                       plot_args))
    
    title_plt <- ggplot() +
      theme_void() + 
      labs(tag = str_wrap(id, 20)) +
      theme(plot.tag.position = c(0.75, 0.5),
            plot.tag = element_text(angle = 0,
                                    size = 18,
                                    face = "bold"))
    
    title_plts <- c(list(title_plt), vector("list", ncols - 1))
    
    if (is.ggplot(plts)) {
      plts <- list(plts)
    }
    plts <- c(title_plts, plts)
    
    if ((n_plts %% ncols) != 0) {
      to_add <- ncols - (n_plts %% ncols)
      plt_lst <- c(plts, vector("list", to_add))
    } else {
      plt_lst <- plts
    }
    
    
    all_plts <- c(all_plts, plt_lst)
    nrows <- ceiling(length(plt_lst) / ncols)
    sub_plot_rel_heights <- c(0.5, rep(1, nrows - 1))
    rel_heights <- c(rel_heights, sub_plot_rel_heights)
  }
  
  
  p <- plot_grid(plotlist = all_plts,
                 nrow = ceiling(length(all_plts) / ncols),
                 ncol = ncols,
                 scale = 0.8,
                 rel_heights = rel_heights)
  
  
  if (!is.null(outfile)){
    save_plot(outfile,
              p,
              nrow = ceiling(length(all_plts) / ncols),
              ncol = ncols,
              base_asp = 1,
              base_height = 2,
              limitsize = FALSE)
  }
  
  
  p
  
}

plot_montage(so, 
             mkr_list, 
             outfile = file.path(fig_dir, "known_marker_umap_summary.pdf"))
```

There is a very clearly a small population of AT1 cells that colocalize with the AT2 cells. *For now these will be grouped with AT2 cells, but we will likely need to do some custom reclustering to segregate these cells.* 

## Compute cell type markers 

```{r}
mkrs <- get_marker_summaries(so, 
                     "coarse_cell_type", 
                     outdir = mkrs_dir, 
                     min_pct = 10,
                     prefix = "cell_types_coarse")

mkrs <- get_marker_summaries(so, 
                     "cell_type_cluster", 
                     outdir = mkrs_dir, 
                     min_pct = 10,
                     prefix = "cell_types_refined")

topx <- mkrs$mkrs %>% 
  filter(logFC > 0, padj < 0.05) %>% 
  mutate(group = factor(group, levels = unique(so$cell_type_cluster))) %>% 
  group_by(group) %>% 
  arrange(padj, desc(logFC)) %>% 
  dplyr::slice(1:10) 
```

Shown below are the top 10 markers of each cell population in a table and heatmap.

```{r, rows.print = 15, columns.print = 15}
topx %>% 
  select(feature, group) %>% 
  group_by(group) %>% 
  mutate(id = row_number()) %>% 
  pivot_wider(names_from = "group", values_from = "feature") %>%
  select(-id)
```


```{r, fig.width = 9, fig.height= 18, warning = FALSE, message = FALSE, fig.cap = "Heatmap of top markers of each cell type (and subcluster)"}
topx_features <- topx %>% 
  pull(feature)

hm <- plot_heatmap(ScaleData(so, features =  topx_features),
             topx_features, 
             group = "cell_type_cluster",
             average = TRUE,
             hmap_options = list(column_names_rot = 90,
                                 col = viridis::magma(256)))

pdf(file.path(fig_dir,
              "cell_types_refined_marker_heatmap.pdf"),
    height = 18,
    width = 9)
draw(hm)
a <- dev.off(); rm(a)

hm
```


## Transgene expression levels per cluster


```{r, fig.cap = " Expression of transgenes across cell types"}
p <- plot_violins(so, 
             "cell_type_cluster", 
             c("tdtomato-sv40",
               "egfp-sv40"),
             split_by = "treatment") +
  labs(x = "",
       y = "Expression")

save_plot(file.path(fig_dir, "transgene_expression_per_cell_type.pdf"),
          p,
          base_asp = 2)

p
```


## Go Terms

```{r}
library(gprofiler2)
library(ggrepel)
# Use top 500 markers 
mkrs <- read_tsv(file.path("markers", "cell_types_refined_cluster_markers.tsv")) %>% 
  filter(logFC > 0,
         pct_in > 10,
         padj < 0.05) %>% 
  group_by(group) %>% 
  dplyr::slice(1:500)
mkrs_split <- split(mkrs, mkrs$group) %>% 
  map(~pull(.x, feature))
go_res <- gost(mkrs_split, 
               ordered_query = TRUE,
               sources = c("GO", 
                          "KEGG", 
                          "REAC", 
                          "TF"))
go_res[["result"]] %>% 
  mutate(parents = unlist(map(parents, ~str_c(.x, collapse = ",")))) %>% 
  write_tsv(., file.path(tbls_dir, "goterms_cell_type_subclusters.tsv"))
go_res[["result"]] %>% 
  mutate(parents = unlist(map(parents, ~str_c(.x, collapse = ","))),
         source = str_replace_all(source, "[[:punct:]]", " ")) %>% 
  dplyr::select(-significant) %>% 
  split(., .$query) %>% 
  openxlsx::write.xlsx(., 
                       file.path(tbls_dir, "goterms_cell_type_subclusters.xlsx"))
```

```{r}
go_res_all <- gost(mkrs_split, 
               ordered_query = TRUE, 
               significant = FALSE,
               sources = c("GO", 
                          "KEGG", 
                          "REAC", 
                          "TF"))
go_res_to_plot <- go_res_all[["result"]] %>% 
  mutate(parents = unlist(map(parents, ~str_c(.x, collapse = ","))),
         source = str_replace_all(source, "[[:punct:]]", " ")) %>% 
  select(-significant) 
```

```{r, fig.height = 9, fig.width = 7}
dbs <- c("GO BP", "GO MF", "GO CC")
top_n_terms <- 5
gterm_hmaps <- map(dbs,
    ~{
  res <- filter(go_res_to_plot, 
                   source == .x,
                   p_value < 0.05) %>% 
    group_by(query) %>% 
    arrange(p_value, .by_group = TRUE) %>% 
    dplyr::slice(1:top_n_terms)
  
  to_plot <- filter(go_res_to_plot,
                    source == .x,
                    term_name %in% res$term_name) %>%
    select(query, p_value, term_name) %>% 
    pivot_wider(names_from = query, values_from = p_value) %>% 
    column_to_rownames("term_name") %>% 
    as.matrix() 
  
  to_plot[is.na(to_plot)] <- 1
  to_plot <- -log10(to_plot)
  hmap <- Heatmap(to_plot, 
          name = "-log10(pvalue)",
          column_title = .x,
          column_order = unique(go_res_to_plot$query),
          row_order = unique(res$term_name),
          col = viridis::magma(256),
          cluster_columns = TRUE, 
          cluster_rows = FALSE, 
          row_names_side = "left")
  hmap
})
walk(gterm_hmaps, draw)
```

## Cell proportions


```{r, fig.cap = "Cell type proportion per treatment"}
props <- get_cell_count_matrix(so, "cell_type_cluster", "sample") %>% 
  rownames_to_column("cell_type") %>% 
  pivot_longer(cols = -cell_type) %>% 
  group_by(name) %>% 
  mutate(percent = 100 * (value / sum(value))) %>% 
  mutate(name = factor(name, levels = c("S1", "S2", "E1", "E2")))

p <- ggplot(props, aes(cell_type, percent)) +
  geom_col(aes(fill = name), position = position_dodge()) +
  scale_fill_manual(name = "",
                    values = discrete_palette_default) +
  theme(axis.text.x = element_text(angle = 90,hjust = 1, vjust = 0.5)) +
  labs(x = "",
       y = "Percent of Cells")

save_plot(file.path(fig_dir, "cell_proportions_per_treatment.pdf"), 
          p, 
          base_asp = 2)
p
```


## Cell cycle

```{r, fig.cap = "UMAP of inferred cell cycle state"}
cc_mouse_genes <- map(cc.genes.updated.2019, get_orthologs)
so <- CellCycleScoring(so, 
                       s.features = cc_mouse_genes$s.genes,
                       g2m.features = cc_mouse_genes$g2m.genes)    

p <- plot_umap(so, "Phase", group = "sample", .cols = palette_OkabeIto)

save_plot(file.path(fig_dir, "umap_of_cell_cycle_phase.pdf"), 
          p, 
          base_asp = 2)
p
```

```{r}
p <- plot_cell_proportions(so, 
                      sample_id = "cell_type_cluster",
                      group_id =  "Phase", 
                      facet_by = "treatment",
                      cols =  palette_OkabeIto) +
  labs(x = "")


save_plot(file.path(fig_dir, "cell_cycle_proportions.pdf"),
          p,
          base_asp = 2)

p
```

## Save data analysis

```{r} 
qsave(so, file.path(obj_dir, "so.qs"), nthreads = 4)

mdata_out <- get_metadata(so)
outname <- file.path("tables", 
                     paste0("metadata_", 
                            format(Sys.Date(), "%Y_%m_%d"),
                            ".tsv.gz"))
write_tsv(mdata_out, outname)

```

